Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-04-26 LICS2021-final.tex
Add the copyright notice
tweaked lics final version

\documentclass[conference,twosided,10pt]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

@@

different syntactic proof systems, based on different sets of inference rules. Just consider the derivations in Figure~\ref{fig:translate}, showing two proofs of the formula $\peirceformula$ and two proofs of the formula $\veedrinkerformula$, one in the sequent calculus (top) and one in a deep inference system (bottom). It is, \emph{a priori}, not clear how to compare them.

@@

}

@@

}
\begin{figure}\vspace{-2ex}\begin{center}\hspace*{-0ex}\begin{math}

@@

This is where \emph{combinatorial proofs} come in. They were
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a syntax-free notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in inductively-generated proofs
and consider the proof as a combinatorial object, more precisely
as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula=\peirceformula$ is shown mid-left in
Fig.\,\ref{fig:translate},
%depicted below-left,
a homomorphism from a coloured graph to a graph labelled with
propositional variables.
%
% \begin{center}\vspace{.5ex}\begin{tikzpicture}\newcommand\radius{2.3}
%   \begin{scope}[shift={(-\radius,0)}]{\peircecp}\end{scope}
%   \begin{scope}[shift={(\radius,0)}]{\drinkerfibcoloured}\end{scope}
% \end{tikzpicture}\vspace{0ex}\end{center}
% %
% 
% \begin{equation*}
%   \scalebox{.7}{$\lkdrinkerproof
% %\qquad
% \ksonedrinkerproof
% $}
% \end{equation*}
% 
% \juihsuan{LK1 and KS1 proofs of Peirce's law and LK1 proof of drinker formula}
% \lutz{can you also do the KS1 for the drinker?}
% \juihsuan{done}

Several authors have illustrated how syntactic proofs in various
proof systems can be translated to propositional combinatorial proofs:
for sequent proofs in~\cite{hughes:invar}, for deep inference proofs
in~\cite{str:fscd17}, for Frege systems in~\cite{str:RR-9048}, and for
tableaux systems and resolution in~\cite{acc:str:18}. This enables a
natural definition of proof identity for propositional logic: two
proofs are \emph{the same}, if they are mapped to the same
combinatorial proof.
%
For example, the left side of Fig.\,\ref{fig:translate} translates
syntactic proofs from sequent calculus and the calculus of structures into
the same combinatorial proofs, witnessing that the two syntactic proofs,
from different systems, are \emph{the same}.

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
``drinker paradox'' $\drinkerformula=\veedrinkerformula$ is shown on the
right of Fig.\,\ref{fig:translate}, %\todo{wrong implies symbol in drinkerformula?}
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is highly unsatisfactory: (1) the soundness argument is extremely
long, intricate and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not
\emph{sequentializable} \cite{girard:87} nor
\emph{full}~\cite{abramsky:jagadeesan:94}. %\dominic{ok to say sequentializable too?}
%
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.

We solve these issues by moving to a deep inference
system. More precisely, we introduce a new proof system, called
$\FOKS$, for first-order logic, that (a) reflects every combinatorial
proof, i.e., there is a surjection from $\FOKS$ proofs to
combinatorial proofs, (b) yields far simpler proofs of
soundness and completeness for combinatorial proofs, and (c) admits
new decomposition theorems establishing a precise correspondence
between certain syntactic inference rules and certain combinatorial
notions.
%
The right side of Fig.\,\ref{fig:translate}
illustrates the surjection in (a), and since the syntactic proofs 
of the two systems both translate to the same combinatorial proof,
they can be considered \emph{the same}.

In general, a \emph{decomposition theorem} provides normal forms of
proofs, separating subsets of inference rules of a proof system. A
prominent example of a decomposition theorem is Herbrand's
theorem~\cite{herbrand:phd}, which allows a separation between the
propositional part and the quantifier part in a first-order
proof~\cite{gentzen:35:II,brunnler:06:herbrand}. Through the advent of
deep inference, new kinds of proof decompositions became possible,
most notably the separation between the linear part of a proof and the
resource management of a proof. It has been shown by
Stra{\ss}burger~\cite{str:07:RTA} that a proof in classical
propositional logic can be decomposed into a proof of multiplicative
linear logic, followed by a proof consisting only of contractions and
weakenings. In this paper we show that the same is possible for
first-order logic.

@@

Fix pairwise disjoint countably infinite sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. Each predicate symbol $p$ has a
\bfit{dual}
$\dual{p}$. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability we often omit parentheses and write simply $ft_1 \dots t_n$ or $pt_1 \dots t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

A formula is \bfit{rectified} if all bound variables are distinct from
one another and from all free variables. Every formula can be
transformed into a logically equivalent rectified form, by simply
renaming the bound variables. If we consider formulas equivalent
modulo $\alpha$-conversion (renaming of bound variables), then the
rectified form of a formula $A$ is uniquely defined, and we denote it
by $\rectif A$.

@@

\bfit{Sequents}, denoted by $\Gamma,\Delta,\ldots$, are finite
multisets of formulas, written as lists, separated by comma. The
\bfit{corresponding formula} of a (non-empty) sequent $\Gamma=A_1,A_2,\ldots,A_n$
is the disjunction of its formulas: $\form\Gamma=A_1\vlor
A_2\vlor\cdots\vlor A_n$. A sequent is \bfit{rectified} iff its
corresponding formula is.
%\dominic{I changed fm$(\Gamma)$ to $\form\Gamma$,
%easier for referee to understand/intuit/remember later in the paper}

In this paper we use the sequent calculus $\FOLK$, shown in
Figure~\ref{fig:LK1}, which is a one-sided variant of Gentzen's
original calculus~\cite{gentzen:35:I} for first-order logic. To
simplify some technicalities later in this paper, we include
the $\mixr$-rule.
%
\begin{thm}
  $\FOLK$ is sound and complete for first-order logic.
\end{thm}
%
\noindent For a proof, see any standard textbook, e.g.~\cite{TS:00}.

@@

@@

A graph $\gG$ is \bfit{labelled} in a set $L$ if each vertex $v\in\vG$
has an element $\labelof v\in L$ associated with it, its \bfit{label}.  A
graph $\gG$ is (partially) \bfit{coloured} if it carries a partial
equivalence relation $\linkingof\gG$ on $\vG$; each equivalence class is a
\bfit{colour}.  A \bfit{vertex renaming} of $\gG=\tuple{\vG,\eG}$
along a bijection $(\hat\cdot)\colon\vG\to\hat\vG$ is the graph
$\hat\gG=\tuple{\hat\vG,\set{\hat v\hat w\mid vw\in\eG}}$, with
colouring and/or labelling inherited (i.e., $\hat v\samecol\hat w$ if
$v\samecol w$, and $\ell(\hat v)=\ell(v)$). Following standard graph
theory, we identify graphs modulo vertex renaming.

@@

In a graph $\gG$, the \bfit{neighbourhood} $N(v)$ of a vertex $v \in\vG$ is defined as
the set $\set{w \mid vw \in\eG}$. A \bfit{module} is a set $M \subseteq \vG$
such that $N(v) \setminus M = N(w) \setminus M$
for all $v, w \in M$. A module $M$ is \bfit{strong} if for every
module $M'$, we have $M' \subseteq M$, $M \subseteq M'$ or $M \cap M' =
\emptyset$. A module is \bfit{proper} if it has two or more vertices.

%% Modules in cographs correspond precisely to the subtrees of the
%% cotrees (the term-trees constructing the graph via $+$ and~$\times$).\todo{Cotree
%% must alternate $+$ and $\times$. Skip this comment completely?}

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. An
atom-labelled vertex is called a \bfit{literal} and a
variable-labelled vertex is called a \bfit{binder}. A binder labelled
with $x$ is called an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.
%
\begin{definition}
A \bfit{first-order graph} or \bfit{fograph} is a logical cograph in which all binders are legal. The
\bfit{binding graph} of a fograph $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}
%
\noindent We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%% For a formula $A$, we can define its
%% associated fograph $\graphof{A}$ inductively by:\todo{duplicate two sentences above}
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    %% \multicolumn{4}{c}{
    %%   \quad\graphof{\ttt}=\single\ttt
    %%   \qquad
    %%   \graphof{\fff}=\single\fff
    %%   \qquad
    %% \graphof{a} = \single a \hbox{\hskip1em(for any atom $a$)} } \\[1.1ex]
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where we
write $\single\alpha$ for a single-vertex  labelled by $\alpha$.

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
  structural induction on $A$.
  %% :
  %% \begin{itemize}
  %%   \item Trivial if $A$ is an atom.
  %%   \item The induction step uses the following fact:
  %%         if $M$ is a module of $\gG$ and $M' \subseteq M$, then $M'$ is a
  %%         module of $\gG$ if and only if it is a module of $\gG[X]$.
  %% \end{itemize}
  %% \todo{}\lutz{maybe mention also fn 2-5 in Dominic's arxiv-paper}
\end{proof}
%%\juihsuan{I'm still looking for a simple proof}
%
\begin{remark}
Note that $\graphof A$ is not necessarily a fograph if $A$ is not
rectified. If $A = (\forall x. px) \vlor (\forall x. qx)$, then
$\graphof A = \single x \ \single px \ \single x \ \single qx$,
the scope of each $x$-binder contains all the vertices, in particular,
the other $x$-binder. On the other hand, there are non-rectified
formulas which are translated to fographs by $\graphof\cdot$. For
example, in the graph of $(\exists x. px) \vlor (\exists x. qx)$,
both $x$-binders are legal, as they are not in each other's
scope:
$\edgepairpic{x}{px}\mkern9mu\edgepairpic{x}{qx}$.
\end{remark}
%
%% For this reason, we call a formula
%% \bfit{clean} if it does not contain subformulas of the form $(\forall
%% x.A)\vlor(\forall x.B)$ or $(\exists x.A)\vlan(\exists x.B)$,
%% \todo{need the four cases as in clear definition in fopws? Consider
%% $(\forall x.A)\vlor(B\vlor (\forall x.C))$. But then Lemma 6 doesn't hold.
%% (fopws proves only that there's a surjection from clear formulas onto fographs,
%% whose kernel is the six formula equalities)} and no
%% $x$-quantified formula occurs as subformula of another $x$-quantified
%% formula. Then we have:
%
%% \begin{lemma}
%%   If $A$ is clean iff\/ $\graphof A$ is a fograph.
%% \end{lemma}
%
%% \begin{proof}
%%   Induction on $A$, using  Theorem~\ref{thm:cograph}.
%% \end{proof}
%
%% Even though for every formula $A$ we can obtain an
%% equivalent clean formula by simply renaming some bound variables, this
%% is not unique up to $\alpha$-conversion, as is the case for
%% rectified formulas.
%
\noindent We define a congruence relation $\fequ$ on formulas by the following equations:
\begin{equation}
  \label{eq:fequ}
  \begin{array}{c@{\hskip1.5em}c}
    A \vlan B \fequ B \vlan A
    &
    (A \vlan B) \vlan C \fequ A \vlan (B \vlan C) 
    \\
    A \vlor B \fequ B \vlor A
    &
    (A \vlor B) \vlor C \fequ A \vlor (B \vlor C)
    \\
    \forall x . \forall y . A \fequ \forall y . \forall x . A
    &
    \forall x . (A \vlor B) \fequ (\forall x . A) \vlor B
    \\
    \exists x . \exists y . A \fequ \exists y . \exists x . A
    &
    \exists x . (A \vlan B) \fequ (\exists x . A) \vlan B
  \end{array}
\end{equation}
where $x$ must not be free in $B$ in the last two equations.
Two formulas $A$ and $B$ are \bfit{equivalent} if $A \fequ B$. We have the
following theorem:
\begin{thm}
  Let $A,B$ be rectified formulas. Then
  $$
  A\fequ B \iff \graphof A =\graphof B
  $$
\end{thm}

\begin{proof}
  Straightforward induction on $A$.
\end{proof}

\begin{example}
  Both $\veedrinkerformula$ and $\variantveedrinkerformula$, which are
  equivalent modulo $\fequ$, have the same (rectified) fograph
  $\drinkergraph$, shown below-left.
  %
  \begin{equation*}
      \drinkergraphpic
      %
      \hspace{8ex}
      %
      \drinkercotreepic
      %
      \hspace{8ex}
      %
      \drinkerbindinggraphpic
  \end{equation*}
  Above-middle we show the \emph{cotree} of the underlying
  cograph (illustrating the idea behind Theorem~\ref{thm:cograph}) and
  above-right is its binding graph $\bD$.
\end{example}

@@

Two atoms are \bfit{pre-dual} if they are not $\ttt$ or $\fff$, and their predicate symbols are dual
(e.g. $p(x, y)$ and $\dual{p}(y, z)$) and two literals are \bfit{pre-dual} if their
labels (atoms) are pre-dual. A \bfit{linked fograph} $\tuple{\gC,\linkingof\gC}$ is a coloured fograph $\gC$ such
that every colour (i.e., equivalence class of~$\linkingof\gC$), called a \bfit{link}, consists of two pre-dual literals, and
every literal is either $\ttt$-labelled or in a link. Hence, in a linked fograph no vertex is labelled~$\fff$.

@@

We state the main results of this paper here,
and prove them in later sections. The
first is routine and expected, but needs to be proved nonetheless:

@@

\begin{lemma}\label{lem:shallow}
  Let $\vlinf{\rr}{}{\Scons B}{\Scons A}$ be an inference rule in
  $\FOMLS$. Then the sequent $\sqn{\cneg A,B}$ is provable in $\FOMLL$.
\end{lemma}
\begin{proof}
  This is a straightforward exercise.
%%  that we leave to the reader. (Note that the
%%$\ax$-rule can be applied to $\sqn{\fff,\ttt}$ in the cases of $\rr = \faD$.)
\end{proof}

@@

\begin{proof}
  Straightforward induction on  $\Sconhole$. (see e.g. \cite{gug:str:01})
\end{proof}

>>>>

\documentclass[conference,10pt]{IEEEtran}
% Some Computer Society conferences also require the compsoc mode option,
% but others use the standard conference format.
%
% If IEEEtran.cls has not been installed into the LaTeX system files,
% manually specify the path to it like:
% \documentclass[conference]{../sty/IEEEtran}

@@

%% \newcommand{\longer}[2]{#1}
%% \newcommand{\shorter}[2]{#2}

%% %\let\choice=\longer
%% \let\choice=\shorter

\IEEEoverridecommandlockouts
%\IEEEpubid{\makebox[\columnwidth]{\emph{[Long version of the LICS 2021 paper, with proofs in the appendix.]}\hfill}
%    \hspace{\columnsep}\makebox[\columnwidth]{ }}
\IEEEpubid{\makebox[\columnwidth]{978-1-6654-4895-6/21/\$31.00~
\copyright2021 IEEE \hfill} \hspace{\columnsep}\makebox[\columnwidth]{ }}

@@

different syntactic proof systems, based on different sets of inference rules.
Consider the derivations in Figure~\ref{fig:translate}, showing two proofs of the
formula $\peirceformula$ and two proofs of the formula $\veedrinkerformula$,
in sequent calculus (top) and in a deep inference system (bottom).
It is, \emph{a priori}, not clear how to compare them.%

@@

}%

@@

}%
\begin{figure}\vspace{-4ex}\begin{center}\hspace*{-0ex}\begin{math}

@@

This is where \emph{combinatorial proofs} come in. They were
introduced by Hughes~\cite{hughes:pws} for classical propositional
logic as a syntax-free notion of proof, and as a potential solution
to Hilbert's 24th problem~\cite{hughes:invar} (see
also~\cite{str:hilbert:24}). The basic idea is to abstract away from
the syntax of the inference rules used in inductively-generated proofs
and consider the proof as a combinatorial object, more precisely
as a special kind of
graph homomorphism. For example, a propositional combinatorial
proof of Peirce's law $\peirceimpliesformula=\peirceformula$ is shown mid-left in
Fig.\,\ref{fig:translate},
%depicted below-left,
a homomorphism from a coloured graph to a graph labelled with
propositional variables.
%

Several authors have illustrated how syntactic proofs in various
proof systems can be translated to propositional combinatorial proofs:
for sequent proofs in~\cite{hughes:invar}, for deep inference proofs
in~\cite{str:fscd17}, for Frege systems in~\cite{str:RR-9048}, and for
tableaux systems and resolution in~\cite{acc:str:18}. This enables a
natural definition of proof identity for propositional logic: two
proofs are \emph{the same} if they are mapped to the same
combinatorial proof.
%
For example, the left side of Fig.\,\ref{fig:translate} translates
syntactic proofs from sequent calculus and the calculus of structures into
the same combinatorial proofs, witnessing that the two syntactic proofs,
from different systems, are \emph{the same}.

@@

In this paper we advance the idea that combinatorial proofs can provide
a notion of proof identity for first-order logic. \emph{First-order
combinatorial proofs} were introduced by Hughes in~\cite{hughes:fopws}.
For example, a first-order combinatorial proof of Smullyan's 
\emph{drinker paradox} \mbox{$\drinkerformula$} $=$ \mbox{$\veedrinkerformula$}
is shown on the right of Fig.\,\ref{fig:translate}, 
a homomorphism from a partially coloured graph to a labelled graph.
However, even though Hughes proves soundness and completeness, the 
proof is unsatisfactory: (1) the soundness argument is long, intricate
and cumbersome, and (2) the completeness proof does not 
allow a syntactic proof to be read back from a combinatorial proof, i.e.,
completeness is not
\emph{sequentializable} \cite{girard:87} nor
\emph{full}~\cite{abramsky:jagadeesan:94}. 
%
A fundamental problem is that not all combinatorial
proofs can be obtained as translations of sequent calculus proofs.

We solve these issues by moving to a deep inference
system. More precisely, we introduce a new proof system,
$\FOKS$, for first-order logic, that (a) reflects every combinatorial
proof, i.e., there is a surjection from $\FOKS$ proofs to
combinatorial proofs, (b) yields simpler proofs of
soundness and completeness for combinatorial proofs, and (c) admits
new decomposition theorems establishing a precise correspondence
between certain syntactic inference rules and certain combinatorial
notions.
%
The right of Fig.\,\ref{fig:translate}
illustrates surjection in (a), and since the syntactic proofs 
in the two systems translate to the same combinatorial proof,
they can be considered \emph{the same}.

In general, a \emph{decomposition theorem} provides normal forms of
proofs, separating subsets of inference rules of a proof system. A
prominent example of a decomposition theorem is Herbrand's
theorem~\cite{herbrand:phd}, which allows a separation between the
propositional part and the quantifier part in a first-order
proof~\cite{gentzen:35:II,brunnler:06:herbrand}. Through the advent of
deep inference, new kinds of proof decompositions became possible,
most notably the separation between the linear part of a proof and the
resource management of a proof. It has been shown by
Stra{\ss}burger~\cite{str:07:RTA} that a proof in classical
propositional logic can be decomposed into a proof of multiplicative
linear logic, followed by a proof consisting only of contractions and
weakenings (see also \cite[\S4]{hughes:invar}).
In this paper we show that the same is possible for
first-order logic.

@@

Fix pairwise disjoint countably infinite sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. Each predicate symbol $p$ has a
\bfit{dual}
$\dual{p}$ with $\dual{\dual p}\mkern6mu{\neq}\mkern5mu\dual p$. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability we often omit parentheses and write $ft_1 \dots t_n$ or $pt_1 \dots t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{c@{\hspace{6ex}}c}
%    \dual{\dual a} = a
%    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
%  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
%  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} \\[.5ex]
%  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Note that $\dual{\dual a} = a$.
We write $A\implies B$ as abbreviation for $\dual A\vlor B$.

A formula is \bfit{rectified} if all bound variables are distinct from
one another and from all free variables. Every formula can be
transformed into a logically equivalent rectified form by
bound variable renaming, e.g.\
%
\mbox{$(px \vee \exists xqx) \wedge \exists x r$} $\mapsto$ \mbox{$(px \vee \exists yqy) \wedge \exists zrz$}.
%
If we consider formulas equivalent
modulo bound variable renaming, the
rectified form of a formula $A$ is uniquely defined, and we denote it
by $\rectif A$.

@@

\bfit{Sequents}, denoted by $\Gamma,\Delta,\ldots$, are finite
multisets of formulas, written as lists, separated by comma. The
\bfit{corresponding formula} of a (non-empty) sequent $\Gamma=A_1,A_2,\ldots,A_n$
is the disjunction of its formulas: $\form\Gamma=A_1\vlor
A_2\vlor\cdots\vlor A_n$. A sequent is \bfit{rectified} iff its
corresponding formula is.

In this paper we use the sequent calculus $\FOLK$, shown in
Figure~\ref{fig:LK1}, which is a one-sided variant of Gentzen's
original calculus~\cite{gentzen:35:I} for first-order logic. To
simplify some technicalities later in this paper, we include
the $\mixr$ rule.
%
\begin{thm}
  $\FOLK$ is sound and complete for first-order logic.
\end{thm}
%
\noindent For a proof, see any standard textbook, e.g.~\cite{TS:00}.

@@

% We recall the definition of \emph{first-order graph} or \emph{fograph}
% \cite[\S3]{hughes:fopws}, used to represent first-order formulas.

@@

A graph $\gG$ is \bfit{labelled} in a set $L$ if each vertex $v\in\vG$
has a \bfit{label} $\labelof v\in L$ associated with it.
A graph $\gG$ is (partially) \bfit{coloured} if it carries a partial
equivalence relation $\linkingof\gG$ on $\vG$; each equivalence class
is a \bfit{colour}.\footnote{In \cite{hughes:fopws}
adjacent vertices are required to have distinct colours, since it
is part of the standard definition of colouring in graph
theory.
However, this condition is redundant (ommitted here), since it is implied
by the preclusion of bimatchings in the next section.}
%Therefore we do not have this additional condition here.}
A \bfit{vertex renaming} of
$\gG=\tuple{\vG,\eG}$ along a bijection
$(\hat\cdot)\colon\vG\to\hat\vG$ is the graph
$\hat\gG=\tuple{\hat\vG,\set{\hat v\hat w\mid vw\in\eG}}$, with
colouring and/or labelling inherited (i.e., $\hat v\samecol\hat w$ if
$v\samecol w$, and $\ell(\hat v)=\ell(v)$). Following standard graph
theory, we identify graphs modulo vertex renaming.

@@

In a graph $\gG$, the \bfit{neighbourhood} $N(v)$ of a vertex $v \in\vG$ is
$\set{w \mkern-2mu\mid\mkern-2mu vw \mkern-2mu\in\mkern-2mu\eG}$. A \bfit{module} is a set $M \mkern-2mu\subseteq\mkern-2mu \vG$
with $N(v) \mkern-1mu\setminus\mkern-1mu M = N(w) \mkern-1mu\setminus\mkern-1mu M$
for all $v, w \in M$. A module $M$ is \bfit{strong} if for every
module $M'$ we have $M' \mkern-2mu\subseteq\mkern-2mu M$, $M \mkern-2mu\subseteq\mkern-2mu M'$ or $M \cap M' =
\emptyset$. A module is \bfit{proper} if it has two or more vertices.

@@

A cograph is \bfit{logical} if every vertex is labelled by either an
atom or variable, and it has at least one atom-labelled vertex. An
atom-labelled vertex is a \bfit{literal} and a
variable-labelled vertex is a \bfit{binder}. A binder labelled
with $x$ is an \bfit{$x$-binder}. The \bfit{scope} of a binder
$b$ is the smallest proper strong module containing $b$.
An \bfit{$x$-literal} is a literal whose
atom contains the variable $x$. An $x$-binder \bfit{binds} every
$x$-literal in its scope.  In a logical cograph $\gG$, a binder $b$ is
\bfit{existential} (resp. \bfit{universal}) if, for every other vertex
$v$ in its scope, we have $bv \in \eG$ (resp. $bv \notin \eG$). An
$x$-binder is \bfit{legal} if its scope contains no other $x$-binder
and at least one literal.
%
\begin{definition}[{\cite[\S3]{hughes:fopws}}]
A \bfit{first-order graph} or \bfit{fograph} $\gG$ is a logical cograph whose binders are all legal. The
\bfit{binding graph} of $\gG$ is the directed graph $\bG=\tuple{\vG,
\set{(b, l) \mid b \text{ binds } l}}$.
\end{definition}
%
\noindent We define a mapping $\graphof\cdot$ from formulas to (labelled)
graphs, inductively as follows:
%
\begin{equation*}
  \begin{array}{r@{\;}l@{\hskip4em}r@{\;}l}
    %% \multicolumn{4}{c}{
    %%   \quad\graphof{\ttt}=\single\ttt
    %%   \qquad
    %%   \graphof{\fff}=\single\fff
    %%   \qquad
    %% \graphof{a} = \single a \hbox{\hskip1em(for any atom $a$)} } \\[1.1ex]
    \graphof{a} &= \single a \rlap{\hskip1em(for any atom $a$)}  \\[.9ex]
    \graphof{A \cor B} &= \graphof{A} + \graphof{B} &
    \graphof{\exists x . A} &= \single x \times \graphof{A} \\[.9ex]
    \graphof{A \cand B} &= \graphof{A} \times \graphof{B} &
    \graphof{\forall x . A} &= \single x + \graphof{A}    
  \end{array}
\end{equation*}
where we
write $\single\alpha$ for a single-vertex  labelled by $\alpha$.

@@

\begin{proof}
  That $\graphof A$ is a logical cograph follows immediately from the definition and Theorem~\ref{thm:cograph}.
  The fact that every binder of $\graphof A$ is legal can be proved by
  structural induction on $A$.
\end{proof}
%
\begin{remark}
Note that $\graphof A$ need not be a fograph if $A$ is not
rectified. If $A = (\forall x. px) \vlor (\forall x. qx)$, then
$\graphof A = \single x \ \single px \ \single x \ \single qx$,
the scope of each $x$-binder contains all the vertices, in particular,
the other $x$-binder. On the other hand, there are non-rectified
formulas which are translated to fographs by $\graphof\cdot$. For
example, in the graph of $(\exists x. px) \vlor (\exists x. qx)$,
both $x$-binders are legal, as they are not in each other's
scope:
$\edgepairpic{x}{px}\mkern9mu\edgepairpic{x}{qx}$.
\end{remark}
%
\noindent We define a congruence relation $\fequ$ on formulas, called \bfit{equivalence},
by the following equations:
\begin{equation}
  \label{eq:fequ}
  \begin{array}{c@{\hskip1.5em}c}
    A \vlan B \fequ B \vlan A
    &
    (A \vlan B) \vlan C \fequ A \vlan (B \vlan C) 
    \\
    A \vlor B \fequ B \vlor A
    &
    (A \vlor B) \vlor C \fequ A \vlor (B \vlor C)
    \\
    \forall x . \forall y . A \fequ \forall y . \forall x . A
    &
    \forall x . (A \vlor B) \fequ (\forall x . A) \vlor B
    \\
    \exists x . \exists y . A \fequ \exists y . \exists x . A
    &
    \exists x . (A \vlan B) \fequ (\exists x . A) \vlan B
  \end{array}
\end{equation}
where $x$ must not be free in $B$ in the last two equations.
%Two formulas $A$ and $B$ are \bfit{equivalent} if $A \fequ B$.
%We have the following theorem:
\begin{thm}[{\cite[\S10]{hughes:fopws}}]
  Let $A,B$ be rectified formulas. Then
  $$
  A\fequ B \iff \graphof A =\graphof B
  $$
\end{thm}

\begin{proof}
  A straightforward structural induction on formulas.
\end{proof}

\begin{example}
  \mbox{$\veedrinkerformula$} $\fequ$ \mbox{$\variantveedrinkerformula$}, and both formulas
  %The formulas \mbox{$\veedrinkerformula$} and \mbox{$\variantveedrinkerformula$}, which are
  %equivalent modulo $\fequ$,
  have the same (rectified) fograph
  $\drinkergraph$, below-left.
  %
  \begin{equation*}
      \drinkergraphpic
      %
      \hspace{8ex}
      %
      \drinkercotreepic
      %
      \hspace{8ex}
      %
      \drinkerbindinggraphpic
  \end{equation*}
  Above-center we show the \emph{cotree} of the underlying
  cograph (illustrating the idea behind Theorem~\ref{thm:cograph}) and
  above-right is its binding graph $\bD$.
\end{example}

@@

Two atoms are \bfit{pre-dual} if they are not $\ttt$ or $\fff$, and their predicate symbols are dual
(e.g.\ $p(x, y)$ and $\dual{p}(y, z)$) and two literals are \bfit{pre-dual} if their
labels (atoms) are pre-dual. A \bfit{linked fograph} $\tuple{\gC,\linkingof\gC}$ is a coloured fograph $\gC$ such
that every colour (i.e., equivalence class of~$\linkingof\gC$), called a \bfit{link}, consists of two pre-dual literals, and
every literal is either $\ttt$-labelled or in a link. Hence, in a linked fograph no vertex is labelled~$\fff$.

@@

We state the main results of this paper here,
and prove them in later sections. The
first is routine and expected, but must be proved nonetheless:

@@

\begin{lemma}\label{lem:shallow}
  Let $\vlinf{\rr}{}{\Scons B}{\Scons A}$ be an inference rule in
  $\FOMLS$. Then the sequent $\sqn{\cneg A,B}$ is provable in $\FOMLL$.
\end{lemma}
\begin{proof}
  A routine exercise.
\end{proof}

@@

\begin{proof}
  A straightforward induction on $\Sconhole$. (see e.g.~\cite{gug:str:01})
\end{proof}
